Image Batching: Batch Ordering Advanced Imaging Tests in the Emergency Department

### 2024-09-09 MgmtSci/MgmtSci.Rmd
getting going

---
title: "To Batch or Not to Batch: Test-Ordering Variability in the ED"
authors:
  - name: Jacob Jameson
    department:  Interfaculty Initiative in Health Policy
    affiliation: Harvard University
    location: Cambridge, MA 
  - name: Soroush Saghafian
    department:  Harvard Kennedy School of Government
    affiliation: Harvard University
    location: Cambridge, MA 
  - name: Robert Huckman
    department:  Harvard Business School
    affiliation: Harvard University
    location: Boston, MA 
  - name: Nicole Hodgson
    department:  Emergency Department
    affiliation: Mayo Clinic
    location: Phoenix, AZ
abstract: |

keywords:

bibliography: references.bib
biblio-style: unsrt
output: rticles::arxiv_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{graphicx}
  - \usepackage{tikz}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
  - \usepackage{caption}
  - \usepackage{adjustbox}
  - \usepackage{multirow}
---

@@

Healthcare delivery, particularly in the emergency department (ED), is a delicate balance that involves ensuring optimal patient outcomes while optimizing resource utilization. Achieving these twin goals requires timely and accurate diagnosis, which in turn enables prompt and appropriate treatment, consequently improving patient prognosis and reducing the likelihood of adverse events. Furthermore, efficient patient discharge from the ED can help alleviate overcrowding, a severe issue with potential consequences including higher complication rates and increased mortality [@bernstein2009].

One important factor that can impact the speed and effectiveness of diagnosis in the ED is the availability and performance of diagnostic tests [@naseim2015]. A variety of diagnostic tests are used in the ED, including laboratory tests, imaging studies, and specialized tests. These tests can provide valuable information about a patient’s condition and help to guide treatment decisions.

A critical question in this context pertains to whether physicians in the ED should batch order diagnostic tests or order them sequentially. This decision essentially represents a tradeoff between reducing patient length of stay and risk of over-testing. Over-testing, or performing unnecessary tests, can lead to increased costs, unnecessary patient anxiety, and potential harm from follow-up of false-positive results [@koch2018]. Conversely, keeping a patient for an extended time to perform all possible tests could lead to ED overcrowding, an issue associated with severe consequences, as mentioned earlier. Instead, what is needed is a reasonable balance between the number of diagnostic tests performed and the total time the patient is kept in the ED before either being admitted or discharged. Several studies have demonstrated that optimizing the ED patient flow process can result in significant improvements [@saghafian2015], however, research surrounding test ordering strategies to improve the patient flow processes remains limited.

In this paper, we use data from two large EDs to quantify the benefits and consequences of batching versus sequentially ordering advanced imaging tests on patient length of stay, re-admission, and total imaging volume. Our empirical strategy exploits random assignment of patients to ED physicians who differ in their propensity to batch-order diagnostic tests. When patients arrive at the ED, they are assigned to a physician based on availability, with no discretion on either side. Thus, patients who arrive at the ED at similar times are randomly assigned to physicians who vary in their willingness to batch order diagnostic tests. We measure physician tendency to batch using a leave-out, residualized measure based on all other patients the physician has seen in the ED in the study period. The tendency measure strongly predicts the ED test batch outcome but is uncorrelated with patient and ED visit characteristics.

With the caveat that other unobserved dimensions of physician care may impact patient outcomes, we employ physician batch tendency as an instrumental variable for having tests batch ordered in the ED to quantify the effect of batching directly. Because of the institutional features of the ED, our research design closely approximates an RCT that assigns patients to batch-ordering or sequential-ordering arm. In the ED, patients have no discretion over choosing providers, and in our specific ED, physicians have discretion over choosing patients, alleviating major selection issues present in other health care settings. Furthermore, physicians exhibit wide variation in practice behavior in batch-ordering, even within the same hospital, while following the same guidelines. Finally, patient-physician interactions in the ED are typically well documented, short, and one-off, constraining physician decision-making to a more limited, better-observed choice set than present in settings such as specialty or primary care.

In sum, exploiting practice variation in ED settings shuts down other (but not all) potential channels besides test batching that are present in other settings, determine length of stay, and impact patient outcomes. This approach allows us to move closer to identifying the causal impact of batch-ordering diagnostic tests on patient outcomes and resource utilization. It is important to note that this paper studies the impact of batch-ordering through a batching decision requiring clinical judgment (within practice norms) rather than through specific hospital policies, differences in adherence to clinical practice guidelines, or substandard care.

# Literature Review

# Theoretical Background and Hypotheses

Emergency departments (EDs) are complex, time-sensitive environments where physicians must make rapid decisions under uncertainty [@asplin2003model; @jarvis2016ed]. Our theoretical model aims to capture the nuanced dynamics of diagnostic decision-making in this high-stakes setting, with a particular focus on the practice of batch ordering imaging tests.

## Diagnostic Decision-Making and Test Ordering in the ED

In the ED, physicians face the challenge of diagnosing and treating patients efficiently while maintaining high-quality care [@croskerry2013cognitive; @johnson2019decision]. Upon a patient's arrival, the physician must decide on a diagnostic strategy, which often involves ordering imaging tests. This decision-making process is influenced by various factors, including the physician's experience, the patient's presenting symptoms, and the ED's current state (e.g., crowding levels) [@cooke2013overview; @bernstein2009effect; @morley2018emergency].

The key distinction in our study is between batch ordering and sequential ordering of imaging tests. In batch ordering, a physician orders multiple tests simultaneously, while in sequential ordering, tests are ordered one at a time, with each subsequent test decision informed by the results of the previous tests. It's crucial to understand that batch ordering does not mean tests are performed simultaneously. Instead, it means that patients enter the queues for multiple tests at once. This distinction is important when considering the implications of batch ordering on ED operations and patient flow [@green2006managing; @batt2019impact].

## Queuing Theory and ED Operations

When a physician batch orders tests, the patient enters multiple queues simultaneously. This can be modeled using parallel queuing systems in operations research [@dai2010load; @zychlinski2019discrete]. While this might seem to expedite the process, it can lead to inefficiencies in resource allocation, as multiple departments (e.g., radiology, laboratory) must allocate resources to a single patient simultaneously, potentially leading to suboptimal resource utilization [@shi2016models; @wang2019data].

Moreover, batch ordering creates an information lag where results from one test cannot inform the decision to proceed with subsequent tests, potentially leading to unnecessary testing [@welch2011overdiagnosed; @o2018overdiagnosis]. This approach can also create blocking effects, where a patient waiting for multiple tests may block other patients who need only one of those tests, potentially increasing overall waiting times [@armony2015heterogeneous; @song2020managing].

## Reasons for Batch Ordering and Theoretical Implications

Physicians may choose to batch order for several reasons. They may believe it will speed up the diagnostic process [@kocher2012effect; @rosen2018quality], or they may be practicing defensive medicine to avoid missing critical diagnoses [@studdert2005defensive; @kanzaria2015emergency]. Batch ordering may also reduce the cognitive load required to make multiple sequential decisions [@croskerry2013cognitive; @nibbelink2018emergency], which can be particularly appealing in busy EDs where physicians are under time pressure [@bernstein2009effect; @morley2018emergency].

From a theoretical perspective, batch ordering presents several implications. It forgoes the opportunity for value of information considerations, where each test result could inform subsequent decisions [@smith2013impact; @akhlaghpour2020value]. Sequential ordering preserves the option to stop testing once a diagnosis is reached, aligning with real options theory in decision-making under uncertainty [@dixit1994investment; @garg2018value]. While batch ordering might seem efficient for individual patients, it could lead to systemic inefficiencies in resource utilization and patient flow [@green2006managing; @batt2019impact].

## Hypotheses Development

Based on this theoretical framework, we develop the following hypotheses:

### Impact on Resource Utilization

*Hypothesis 1: Batch ordering of diagnostic tests will lead to increased overall test utilization compared to sequential ordering.* This hypothesis is grounded in the idea that batch ordering bypasses the opportunity to use information from initial tests to inform the necessity of subsequent tests, potentially leading to unnecessary testing [@welch2011overdiagnosed; @o2018overdiagnosis].

### Effect on Length of Stay

*Hypothesis 2: Batch ordering will have a significant impact on patient length of stay (LOS) in the ED, but the direction of this effect is ambiguous.* While batch ordering might reduce LOS by eliminating waiting times between test decisions [@kocher2012effect; @rosen2018quality], it could also increase LOS due to potential queuing inefficiencies and the time required to perform and interpret additional tests [@forster2003effect; @armony2015heterogeneous; @song2020managing].

### Quality of Care and Return Visits

*Hypothesis 3: Batch ordering will influence the rate of 72-hour return visits with admission, but the direction of this effect is uncertain.* Comprehensive initial testing might reduce missed diagnoses and thus return visits [@pines2010national; @sabbatini2016overzealous]. However, it could also lead to false positives and unnecessary follow-ups, potentially increasing return visits [@welch2011overdiagnosed; @o2018overdiagnosis].

### Hypothesis 4: Heterogeneity Across Presenting Complaints

*Hypothesis 4: The effects of batch ordering on resource utilization, length of stay, and quality of care will vary significantly across different presenting complaints.* This hypothesis is based on the understanding that different clinical presentations have varying levels of diagnostic uncertainty and risk [@cooke2013overview; @johnson2019decision], which may influence the appropriateness and impact of batch ordering.

By testing these hypotheses, we aim to provide a comprehensive understanding of the impacts of batch ordering in the ED, informing both clinical practice and operational policy. Our empirical strategy, detailed in the following sections, is designed to rigorously test these hypotheses and uncover the causal effects of batch ordering on ED performance and patient outcomes.

# Data and Methods

Our study uses data from two large U.S. emergency departments (EDs): the Mayo Clinic of Arizona and Massachusetts General Hospital (MGH). The MGH dataset, which includes 129,489 patient encounters from November 10, 2021 through December 10, 2022, provides a robust sample for validating the generalizability of our findings. However, our primary analysis focuses on the Mayo Clinic data due to its unique feature of random patient-physician assignment, which allows for stronger causal inference.

## Mayo Clinic Data

@@

\begin{table}[ht]
\centering
\caption{Summary Statistics of Mayo Clinic Emergency Department Encounters}
\label{tab:summary_stats}
\begin{threeparttable}
\begin{tabular}{p{10cm}ccccc}
\toprule
Variable &  & {Mean} & {Q1} & {Median} & {Q3} \\
\midrule
\textbf{Emergency Department Characteristics} & & & & & \\
Total Patients & & {48,854} & {-} & {-} & {-} \\
Patients Admitted & & {18.7\%} & {-} & {-} & {-} \\
Patients Revisited within 72 Hours & & {3.81\%} & {-} & {-} & {-} \\
Patients with IV Fluids & & {35.7\%} & {-} & {-} & {-} \\
Patients with IV Meds & & {17.9\%} & {-} & {-} & {-} \\
Patients in ED & & {24.1} & {-} & {-} & {-} \\
Tachycardic & & {19.2\%} & {-} & {-} & {-} \\
Tachypneic & & {8.82\%} & {-} & {-} & {-} \\
Febrile & & {2.16\%} & {-} & {-} & {-} \\
Hypotensive & & {1.42\%} & {-} & {-} & {-} \\
ESI & & {2.81} & {-} & {-} & {-} \\
Time from Arrival to Triage (mins) & & {8.03} & {4} & {6} & {10} \\
Time from Triage to First Contact (mins) & & {80.2} & {11} & {29} & {61} \\
Average ED LOS (min) & & {246} & {-} & {-} & {-} \\
\midrule
\textbf{Patient Demographics} & & & & & \\
Percent Male & & {46.5\%} & {-} & {-} & {-} \\
Race: White & & {88.4\%} & {-} & {-} & {-} \\
Race: Black & & {4.17\%} & {-} & {-} & {-} \\
Race: Asian & & {3.05\%} & {-} & {-} & {-} \\
Gender: Female & & {53.5\%} & {-} & {-} & {-} \\
Arrival Age & & {57.7} & {43} & {61} & {74} \\
\midrule
\textbf{Diagnostic Tests and Outcomes} & & & & & \\
X-Rays Performed & & {43.3\%} & {-} & {-} & {-} \\
Ultrasounds Performed & & {11.3\%} & {-} & {-} & {-} \\
CTs Performed & & {35.5\%} & {-} & {-} & {-} \\
Labs Ordered & & {73.7\%} & {-} & {-} & {-} \\
Patients Discharged & & {66.8\%} & {-} & {-} & {-} \\
Patients Admitted & & {18.7\%} & {-} & {-} & {-} \\
Contrast CT Performed & & {17.7\%} & {-} & {-} & {-} \\
Time to Result: X-Ray (mins) & & {67.2} & {36} & {54} & {79} \\
Time to Result: Ultrasound (mins) & & {165} & {71} & {101} & {150} \\
Time to Result: Contrast CT (mins) & & {142} & {86} & {115} & {153} \\
Time to Result: Non-Contrast CT (mins) & & {89.7} & {50} & {70} & {102} \\
Time to Result: Lab (mins) & & {46.1} & {25} & {35} & {49} \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{This table reports summary statistics for the baseline sample of emergency department visits during the study period described in the text. Vital signs were categorized as follows: tachycardia (pulse more significant than $100$), tachypnea (respiratory rate greater than $20$), fever (temperature greater than $38^\circ C$), and hypotension (systolic blood pressure less than $90$).}
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

\begin{table}[htbp]
    \centering
    \caption{Balancing Test: Wald Test for Equality of Means}
    \label{tab:wald_test}
    \begin{tabular}{p{10cm}cccc}
        \toprule
        Chief Complaints & Frequency & F-Statistic & $Pr(> F)$ \\
        \midrule
        Abdominal Complaints & 6,232 & 2.587 & 0.108 \\ 
        Back or Flank Pain & 2,552 & 1.637 & 0.201 \\ 
        Chest Pain & 3,525 & 0.407 & 0.524 \\ 
        Extremity Complaints & 5,265 & 1.847 & 0.174 \\ 
        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 2,381 & 0.023 & 0.880 \\ 
        Gastrointestinal Issues & 3,323 & 0.105 & 0.746 \\ 
        Neurological Issue & 3,495 & 0.135 & 0.713 \\ 
        Shortness of Breath & 2,966 & 1.324 & 0.250 \\ 
        Skin Complaints & 2,178 & 0.383 & 0.536 \\ 
        Upper Respiratory Symptoms & 1917 & 0.017 & 0.896 \\ 
        \midrule
        Emergency Severity & Frequency & F-Statistic & $Pr(> F)$ \\
        \midrule
        ESI 1 or 2 & 13,914 & 0.011 & 0.915 \\ 
        ESI 3, 4, or 5 & 29,386 & 0.010 & 0.921 \\ 
        \bottomrule
    \end{tabular}
\begin{tablenotes}
\small
\item \textit{Table \ref{tab:wald_test} reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset. A balanced distribution implies that complaints and severity are evenly distributed across providers, which we expect to be the case due to randomization. The Wald F-statistic and p-value are reported. Robust standard errors (type HC1) were used to account for potential heteroscedasticity in the data.}
\end{tablenotes}
\end{table}

@@

## Sample Construction

Our research design focuses on adults who visit the Mayo Clinic of Arizona ED. We observe $48,854$ such visits during the study period. To improve power, we drop encounters with rare chief complaints ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $5$ percent of the time. Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem if we included them. Examples of complaints dropped include Upper Respiratory Symptoms and Urinary Complaints. Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior. While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories as we do. In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $15,821$ encounters involving full-time physicians who treat over 500 ED cases per year.

## Variable Definitions

Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch-ordered at their ED encounter. While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice. We define "batching" in line with standard emergency medicine practices. Batching occurs when a physician simultaneously orders a comprehensive set of diagnostic tests, typically covering a broad range of potential diagnoses. This contrasts with sequential ordering, where tests are ordered sequentially based on the information obtained from each test as needed.

We operationalize batching as occurring when multiple diagnostic imaging tests are ordered within a 5-minute window. In Section [...], sensitivity analyses on this cutoff point showed that our results are robust to this definition. Each imaging test (e.g., X-ray, CT scan, Ultrasound) is considered a separate, distinct test for our study. Therefore, a batch in our study consists of two or more distinct imaging tests.

Our dependent variables are as follows:

*(a) ED length of stay (LOS).*-- ED-LOS is a critical measure of efficiency and patient throughput in emergency care settings. It is defined as the duration from a patient's arrival to the ED until their departure, whether by discharge or admission to the hospital.

*(b) Number of Unique Imaging Tests*-- Resource utilization in the ED typically refers to the extent of medical services and interventions a patient receives. In this study, we quantify resource utilization by the total number of distinct imaging tests ordered per patient during their ED stay. This encompasses both initial and subsequent tests.

*(c) 72 Hour Return with Admission*-- This is a binary variable indicating whether the patient returns to the ED within 72 hours of their initial visit and is admitted to the hospital. This outcome is used to assess the quality of care provided during the initial ED visit.

## Empirical Strategy

Our empirical strategy closely follows the literature that relies on quasi-random assignment of agents to cases, often referred to as the “judges design.” Papers in this literature typically exploit variation in the sentencing leniency of judges who work in the same court. Similarly, we explore batching variation across physicians who work in the same emergency department. In its reduced form, under the assumption of quasi-random assignment, this approach allows researchers to identify the causal effect of being assigned to different types of physicians. Under additional assumptions, an instrumental variable approach identifies the causal effect of a given medical decision. We employ both approaches and lay out their details in the next subsections.

### Batch Tendency Construction

@@

Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour return with admission, and number of imaging tests ordered for the patient, and $X_i$ is the same as in the reduced-form approach. $Batched_i$ variable suffers from potential endogeneity concerns. For example, injury severity may be unobserved and correlated with need to run multiple tests, length of stay, and return with admission likelihood. Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$. We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.

## Identifying Assumptions

The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, since assignment of patients to ED physicians is random, conditional on seasonality and shift (``conditional independence"). The residualization in equation (1) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch.

@@

# Results {#sec:4}

## Reduced-Form Results

In this section, we explore the causal influence of physician batch tendency on patient outcomes and resource utilization in the emergency department. We posit that while batch tendency directly influences the practice of batch ordering tests, both batch tendency and batch ordering are concurrently influenced by a physician's testing inclination. Given that testing inclination directly affects primary outcomes, we include it as a control variable in our regression models to mitigate its confounding effects.

>>>>

---
journal: mnsc
title: "Image Batching: Batch Ordering Advanced Imaging Tests in the Emergency Department"
shorttitle: "Image Batching"
authors:
- familyname: Jameson
  othernames: Jacob
  affiliation: Harvard Kennedy School, Harvard University
  email: jacobjameson@g.harvard.edu
- familyname: Soroush
  othernames: Saghafian
  affiliation: Harvard Kennedy School, Harvard University
- familyname: Huckman
  othernames:  Robert
  affiliation: Harvard Business School, Harvard University
- familyname: Hodgson
  othernames:  Nicole
  affiliation: Mayo Clinic, Emergency Department
keywords:
  - 3 to 6 keywords
  - that do not appear in the title
abstract: |
  Abstract goes here
acknowledgment: |
  The authors would like to thank ...
bibliography: references.bib
blind: false
output: rticles::informs_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{graphicx}
  - \usepackage{tikz}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
  - \usepackage{caption}
  - \usepackage{adjustbox}
  - \usepackage{multirow}
---

@@

Emergency departments (EDs) are high-pressure environments which globally face escalating challenges such as overcrowding, resource limitations, and increasing patient volumes [@mostafa2024strategies, @Sorup2013evaluation]. These challenges can often be exacerbated by the need to provide timely and accurate diagnoses to patients, which often requires the use of advanced imaging tests such as computed tomography (CT) scans, ultrasound, X-ray, and magnetic resonance imaging (MRI) [@waheed2022impact]. The management of diagnostic testing is a critical component of operational performance in EDs, where timely and accurate diagnoses can significantly impact patient outcomes [@committee2015improving]. In many EDs, however, physicians have considerable discretion over how they order these tests, particularly with regard to advanced imaging [@valtchinov2019use]. Yet little is known about the factors driving physicians' decisions to exercise such discretion and how test ordering should be managed when this discretion exists. In this paper, we explore the causal effects of batch ordering advanced imaging tests, a practice in which physicians order multiple tests simultaneously, on operational performance and patient outcomes in the ED.

Worker discretion can improve system performance, but can sometimes enable workers to "choose the 'wrong' task (operationally)" [@vanDonselaar2010ordering]. We consider the physician's decision to order imaging tests for their patient as an optimization problem, where the physician must balance the trade-offs between the benefits of ordering multiple tests simultaneously at the start of the patient encounter, which may expedite the diagnostic process if the tests are necessary for diagnosis and disposition, and the costs of ordering multiple tests simultaneously, which may increase the total time spent in the ED and the resource utilization of the department. Batching stands in contrast to the more common practice of ordering tests sequentially, where physicians order one test at a time, review the results, and then decide whether to order additional tests. The decision to batch order imaging tests is a form of discretion that physicians can exercise in the ED.

We address two research questions. First, what are the drivers of physician variation in the decision to batch order imaging tests? Second, what are the performance implications of batch ordering imaging tests? To identify the drivers of physician variation in the ordering of imaging tests, we consider the circumstances under which physicians are more likely to batch order imaging tests. We posit that the ability of physicians to identify an alternative test ordering strategy that they perceive as superior to the standard strategy will depend on the characteristics of the physician as well as the characteristics of the patient. With respect to the physician, we examine the role of physician experience. As for patient characteristics, we examine whether a patient has a complex chief complaint or a high acuity level, which may require additional diagnostic testing.

We investigate these questions using operational data from two large EDs in the United States. We focus on imaging tests, which are commonly ordered in the ED and are associated with significant resource utilization. We use a unique dataset that contains detailed information on the timing of test orders, the timing of test results, and the timing of patient disposition. We exploit the random assignment of patients to physicians to identify the causal effects of batch ordering imaging tests on operational performance and patient outcomes. We find that physicians deviate from the standard test ordering strategy 42% of the time. We show that physicians are more likely to batch order imaging tests when they are more experienced, when there is an opportunity to follow a superior test ordering strategy by deviating, and when there is an opportunity to batch by deviating. When physicians deviate from the standard test ordering strategy, their average time to disposition increases by 13%. Other performance dimensions, including resource utilization and patient outcomes, are mostly unaffected. Our calculations suggest that forgoing deviations would have led to faster time to disposition, which could have saved 2,494 hours per year, increasing annual profits by 3%.

We also find that the

# Related Literature

# Batch Ordering Advanced Imaging

## Drivers of Image Batching

## Implications for Emergency Department Operations and Patient Outcomes

## Hetereogeneity in Orderging Strategy

# Setting, Data, and Models

## Empirical Setting

Our study uses data from two large U.S. emergency departments (EDs): the Mayo Clinic of Arizona and Massachusetts General Hospital (MGH). The MGH dataset, which includes 129,489 patient encounters from November 10, 2021 through December 10, 2022, provides a robust sample for validating the generalizability of our findings. However, our primary analysis focuses on the Mayo Clinic data due to its unique feature of random patient-physician assignment, which allows for stronger causal inference. The data contain information on the timing of test orders, the timing of test results, and the timing of patient disposition, among various other important triage metrics and demographic features. We focus on imaging tests (x-rays, contrast CT, non-contrast CT, ultrasound) because unlike laboratory tests, imaging tests cannot be run simultaneously. Therefore, the operational implications of batch ordering imaging tests are more pronounced.

## Data

@@

\begin{table}[ht]
\centering
\caption{Summary Statistics of Mayo Clinic Emergency Department Encounters}
\label{tab:summary_stats}
\begin{threeparttable}
\begin{tabular}{lccccc}
\toprule
Variable &  & {Mean} & {Q1} & {Median} & {Q3} \\
\midrule
\textbf{Emergency Department Characteristics} & & & & & \\
Total Patients & & {48,854} & {-} & {-} & {-} \\
Patients Admitted & & {18.7\%} & {-} & {-} & {-} \\
Patients Revisited within 72 Hours & & {3.81\%} & {-} & {-} & {-} \\
Patients with IV Fluids & & {35.7\%} & {-} & {-} & {-} \\
Patients with IV Meds & & {17.9\%} & {-} & {-} & {-} \\
Patients in ED & & {24.1} & {-} & {-} & {-} \\
Tachycardic & & {19.2\%} & {-} & {-} & {-} \\
Tachypneic & & {8.82\%} & {-} & {-} & {-} \\
Febrile & & {2.16\%} & {-} & {-} & {-} \\
Hypotensive & & {1.42\%} & {-} & {-} & {-} \\
ESI & & {2.81} & {-} & {-} & {-} \\
Time from Arrival to Triage (mins) & & {8.03} & {4} & {6} & {10} \\
Time from Triage to First Contact (mins) & & {80.2} & {11} & {29} & {61} \\
Average ED LOS (min) & & {246} & {-} & {-} & {-} \\
\midrule
\textbf{Patient Demographics} & & & & & \\
Percent Male & & {46.5\%} & {-} & {-} & {-} \\
Race: White & & {88.4\%} & {-} & {-} & {-} \\
Race: Black & & {4.17\%} & {-} & {-} & {-} \\
Race: Asian & & {3.05\%} & {-} & {-} & {-} \\
Gender: Female & & {53.5\%} & {-} & {-} & {-} \\
Arrival Age & & {57.7} & {43} & {61} & {74} \\
\midrule
\textbf{Diagnostic Tests and Outcomes} & & & & & \\
X-Rays Performed & & {43.3\%} & {-} & {-} & {-} \\
Ultrasounds Performed & & {11.3\%} & {-} & {-} & {-} \\
CTs Performed & & {35.5\%} & {-} & {-} & {-} \\
Labs Ordered & & {73.7\%} & {-} & {-} & {-} \\
Patients Discharged & & {66.8\%} & {-} & {-} & {-} \\
Patients Admitted & & {18.7\%} & {-} & {-} & {-} \\
Contrast CT Performed & & {17.7\%} & {-} & {-} & {-} \\
Time to Result: X-Ray (mins) & & {67.2} & {36} & {54} & {79} \\
Time to Result: Ultrasound (mins) & & {165} & {71} & {101} & {150} \\
Time to Result: Contrast CT (mins) & & {142} & {86} & {115} & {153} \\
Time to Result: Non-Contrast CT (mins) & & {89.7} & {50} & {70} & {102} \\
Time to Result: Lab (mins) & & {46.1} & {25} & {35} & {49} \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{This table reports summary statistics for the baseline sample of emergency department visits during the study period described in the text. Vital signs were categorized as follows: tachycardia (pulse more significant than $100$), tachypnea (respiratory rate greater than $20$), fever (temperature greater than $38^\circ C$), and hypotension (systolic blood pressure less than $90$).}
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

\begin{table}[ht]
    \centering
    \caption{Balancing Test: Wald Test for Equality of Means}
    \label{tab:wald_test}
    \begin{threeparttable}
    \begin{tabular}{lcccc}
        \toprule
        Chief Complaints & Frequency & F-Statistic & $Pr(> F)$ \\
        \midrule
        Abdominal Complaints & 6,232 & 2.587 & 0.108 \\ 
        Back or Flank Pain & 2,552 & 1.637 & 0.201 \\ 
        Chest Pain & 3,525 & 0.407 & 0.524 \\ 
        Extremity Complaints & 5,265 & 1.847 & 0.174 \\ 
        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 2,381 & 0.023 & 0.880 \\ 
        Gastrointestinal Issues & 3,323 & 0.105 & 0.746 \\ 
        Neurological Issue & 3,495 & 0.135 & 0.713 \\ 
        Shortness of Breath & 2,966 & 1.324 & 0.250 \\ 
        Skin Complaints & 2,178 & 0.383 & 0.536 \\ 
        Upper Respiratory Symptoms & 1917 & 0.017 & 0.896 \\ 
        \midrule
        Emergency Severity & Frequency & F-Statistic & $Pr(> F)$ \\
        \midrule
        ESI 1 or 2 & 13,914 & 0.011 & 0.915 \\ 
        ESI 3, 4, or 5 & 29,386 & 0.010 & 0.921 \\ 
        \bottomrule
    \end{tabular}
\begin{tablenotes}
\item \textit{Table \ref{tab:wald_test} reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset. A balanced distribution implies that complaints and severity are evenly distributed across providers, which we expect to be the case due to randomization. The Wald F-statistic and p-value are reported. Robust standard errors (type HC1) were used to account for potential heteroscedasticity in the data.}
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

Our research design focuses on adults who visit the Mayo Clinic of Arizona ED. We observe $48,854$ such visits during the study period. To improve power, we drop encounters with rare "reasons for visit" (RVF) ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $5\%$ of the time. Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem if we included them. Examples of complaints dropped include Skin Complaints and Urinary Complaints, as well as other complaints where multiple modalities of imaging is unlikely to occur. Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior. While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories as we do. In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $15,821$ encounters involving full-time physicians who treat over 500 ED cases per year.

### Instrumental Variable

\
Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch ordered during their ED encounter. We define "batching" in line with standard emergency medicine practices. Batching occurs when a physician simultaneously orders a comprehensive set of diagnostic tests, typically covering a broad range of potential diagnoses. This contrasts with sequential ordering, where tests are ordered sequentially based on the information obtained from each test as needed.

We operationalize batching as occurring when multiple diagnostic imaging tests are ordered within a 5-minute window. In Section 5.6, sensitivity analyses on this cutoff point showed that our results are robust to this definition. Each imaging test (e.g., X-ray, Contrast CT scan, Non-Contrast CT scan,  Ultrasound) is considered a separate, distinct test for our study. Therefore, a batch in our study consists of two or more distinct imaging tests.

### Dependent Variables

\
\textit{Logarithm of Emergency Department Length of Stay (LOS)}: ED LOS is a key performance metric for ED operations. It is defined as the time from patient arrival to the ED to the time of discharge from the ED. We use this metric to evaluate the impact of batch ordering on ED efficiency. This variable is measured in minutes and in our sample is right-skewed. We log-transform this variable to normalize its distribution to meet the assumptions of linear regression.

\textit{Logarithm of Time to Last Result}: Time to result is defined as the time from when a diagnostic test is ordered to when the result is available in the electronic medical record. The time stamps are at the minute level; that is, for each case, we know the minute in which a test is ordered by the physician and the minute in which the result is available. In order to fully capture the impact of batch ordering on the efficiency of diagnostic testing, we define time to last result (TIME TO RESULT) as the time from when the first image test is ordered to the the time when the last image test result is available. Time to result as a dependent variable is an understudied metric in the literature, and we believe it may capture a less noisy signal of the efficiency of diagnostic testing than LOS in certain circumstances. This variable is also right-skewed in our sample, so we log-transform it to normalize its distribution.

\textit{Total Imaging Tests Ordered}: Total imaging tests ordered (NUM_IMAGES) is a count of the number of distinct imaging tests ordered for a patient during the ED encounter. We use this variable to evaluate the impact of batch ordering on the number of imaging tests ordered.

\textit{72 Hour Return with Admission}: 72 Hour Return with Admission (72HRA) is a binary variable indicating whether a patient returns to the ED within 72 hours of their initial visit and is admitted to the hospital. This variable is used to evaluate the impact of batch ordering on patient outcomes.

## Identification Strategy

Our empirical strategy closely follows the literature that relies on quasi-random assignment of agents to cases, often referred to as the “judges design.” Papers in this literature typically exploit variation in the sentencing leniency of judges who work in the same court. Similarly, we explore batching variation across physicians who work in the same emergency department. In its reduced form, under the assumption of quasi-random assignment, this approach allows researchers to identify the causal effect of being assigned to different types of physicians (i.e. physiancs with a higher or lower tendency to batch order imaging). Under additional assumptions, an instrumental variable approach identifies the causal effect of a given clincal decision.

@@

Where $Y_i$ represents our main outcomes of interest: ED LOS, TIME TO RESULT, NUM IMAGES, 72HRA, and $X_i$ is the same as in the reduced-form approach. $Batched_i$ variable suffers from potential endogeneity concerns. For example, injury severity may be unobserved and correlated with need to run multiple tests, length of stay, and return with admission likelihood. Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$. We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.

### Identifying Assumptions

\
The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, since assignment of patients to ED physicians is random, conditional on seasonality and shift (``conditional independence"). The residualization in equation (1) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch.

@@

## Econometric Models

# Results and Discussion
## Determenants of Image Batching

To investigate the drivers of batching behavior, we estimate the following regression model:

\small
\begin{equation}
Batched_{i,j} = \beta_0 + \beta_1 EXPERIENCE_{i,j} + \beta_2 SEXM_{i,j} + \beta_3 PATIENTS\_TBS + \beta_4 OCCUPANCY + X_{i,j} + \epsilon_i
\end{equation}

\normalsize
Where $Batched_{i,j}$ is a binary measure of whether physician $j$ batched tests for patient $i$. $EXPERIENCE_{i,j}$ is the number of years since the physician's residency graduation. $SEXM_{i,j}$ is a binary variable indicating the sex of the physician is male. $PATIENTS\_TBS$ is the number of remaining patients to be seen by the physician, excluding patient $i$, at the time of the patient's visit. $OCCUPANCY$ is the number of patients in the ED at the time of the arrival of the patient's visit. $X_{i,j}$ is the vectors of patient covariates described in the previous section. We cluster robust standard errors at the physician level.
